How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-04-18 2.10-sortware.Rmd
Add files via upload
# some changes after reading Parts I and II

## What software should I use to write my PhD? {#software}

Thirty years ago, when I started to use computers as an undergraduate, the word processing standard was a piece of software called WordPerfect. Back then it seemed inconceivable that anyone would want to use anything else, or that WordPerfect could lose its grip on the market. WordPerfect became a victim of the Microsoft revolution that standardised most PC operating systems (think Windows) and software using Microsoft Office in the mid-1990s. The chances are that today your institution will still have subscriptions to Microsoft Office making it free for you to use. But given that changes in software can happen so fast and appear to change with increasing frequency, should you use Microsoft's Word to write your PhD?

The simple answer is that you should use whatever you, and your advisor, feel most comfortable with. You will face a lot of challenges during your PhD including learning to write with lots of different kinds of software. Each one you learn to use will require a steep learning curve and that requires time. If there's any good time to start with a new word processing package it will be before you start your PhD. If you do decide to go with a new writing package then my suggestion would be to choose open source software like latex or R markdown. More on these later.

If you feel comfortable with a word processing package with a nice GUI, like Microsoft Word, then go ahead and use it. My best suggestion is that you do not save your files in their default *.docx format. Instead make sure that all of your files are saved as *.rtf or html. This should make little difference to your opening and editing but will mean that all of your texts are forward compatible with other kinds of software. It will reduce your file size and reduce the chances of file corruption: i.e. when part of the file malfunctions and you lose the contents. That’s right, in addition to loss of data by back up deficiencies [see Part III](data.html), you also have the risk of having corrupt files that will also mean you lose all of the contents. Even if you always use Word and never use any other file format, do yourself a favour and save each version as a *.rtf or *.txt file as well. Just like a back up, this means that you won’t have lost all of the text if a file goes corrupt. And yes, I do know people who have multiple backup versions of corrupt files: it’s happened to me.

The real problem with word processing packages like Microsoft Word is when files get very large as they tend to do during a PhD. Once you start putting chapters together especially if they include embedded images tables and the like you will find that many word processors start misbehaving. This can be especially frustrating in the closing phases of your PhD when you're trying to submit it. Especially if there's a tight deadline. However this problem is easily overcome by chunking your thesis into documents equivalent to chapters and then merging PDFs from each chapter.

### Alternatives to conventional word processors

Many academics are turning to latex in order to write their manuscripts as this gives a lot of possibility to insert links, formulae, references, images, etc in a very simple way. Once you are ready to print a latex document it can be shared very easily as a PDF.

Be aware that the problems may come not with what you want but with how your advisor and or co-authors are comfortable with. I find that commenting on a PDF takes more time than doing the same on a word document. I don't mind working with latex but I know some advisors are very adverse to having to use this platform. Obviously in order to get the feedback from your advisor you are going to need to use a platform that they are also comfortable with. So please consult them on what they are happy to use. You may also need to instruct them about how to make comments. For example track changes is a function that a lot of people like to use but it doesn't translate back into other platforms seamlessly.

As with many things in this book I urge you to keep an open mind with respect to software. For example if your lab is progressive and already experimenting with new and flexible software platforms then I suggest that you learn with them. The future of computing is certainly likely to be open-source software and so the sooner that you become more familiar with this the easier you will find it to adapt in future.

Another consideration is how your word processing package is going to work with referencing and citations. Again your lab may already use some software that you want to integrate with your word processor.

@@

 - Reproducibility
 - Dynamic Content (include statistical code and graphics inside your thesis)
 - Focus on content (not formatting)
 - Numerous output types
 - Independent reference manager (via .bib file) and style files (via .csl file)
 - Free software and community support

Importantly, what you want to try to avoid is conducting a major switch between software right at the end of your PhD. Try making such changes when you have more time, like when you’ve just finished.

### Naming versions of files

Good file management is an important aspect of a PhD. I like my students to give each chapter of their thesis a nickname or short running title that we both agree on to name the files. An example might be “Swimming_performance_V1.rft”. The reason why I don’t like “Chapter_1.rtf” as a filename is that every student has a chapter 1, so I can’t distinguish between them. In addition, I’m not good at remembering which chapter is chapter 1 and which is chapter 2. Sometimes, towards the end of a thesis, the chapters get switched around anyway, causing even more confusion. Notice that I don’t like having spaces in filenames. This is a hangover from old days of MS-DOS, but is consistently useful today when using R and other coding software. However, some software platforms react oddly with underscore (_), and so you might need to work with a hyphen instead of a space (-).

@@

While you and your advisor can adopt any naming convention you like, you and they need to instantly be able to distinguish between a new version and one that they have already commented on by the file name.

### How many versions of each file should you keep?

I would suggest keeping at least one older version of each file to minimise the possibility of losing the content through corrupt files, but you should also be keeping those *.rtf or *.txt files as well, just in case. Because of my early experience with file corruption, I started to keep all versions of files. This is unnecessary, and I’ve never needed to go back to anything but the last version. Nevertheless, text files take up very little space and it’s not a problem to keep more versions if you want to.

>>>>

# What software should I use to write my PhD? {#software}

Thirty years ago, when I started to use computers as an undergraduate, the word processing standard was a piece of software called WordPerfect. Back then it seemed inconceivable that anyone would want to use anything else, or that WordPerfect could lose its grip on the market. WordPerfect became a victim of the Microsoft revolution that standardised most PC operating systems (think Windows) and software using Microsoft Office in the mid-1990s. The chances are that today your institution will still have subscriptions to Microsoft Office making it free for you to use. But given that changes in software can happen so fast and appear to change with increasing frequency, should you use Microsoft's Word, or another piece of software to write your PhD?

The simple answer is that you should use whatever you, and your advisor, feel most comfortable with. You will face a lot of challenges during your PhD including learning to write with lots of different kinds of software. Each one you learn to use will require a steep learning curve, and that will require your time. If there's any good time to start with a new word processing package it will be before you start your PhD. If you do decide to go with a new writing package then my suggestion would be to choose open source software like [Rmarkdown @xie2018r]. More on these later.

If you feel comfortable with a word processing package with a nice GUI, like Microsoft Word, then go ahead and use it. My best suggestion is that you do not save your files in their default .docx format. Instead make sure that all of your files are saved as .rtf or html. This should make little difference to your opening and editing but will mean that all of your texts are forward compatible with other kinds of software. It will reduce your file size and reduce the chances of file corruption: i.e. when part of the file malfunctions and you lose the contents. That’s right, in addition to loss of data by back up deficiencies [see Part III](data.html), you also have the risk of having corrupt files that will also mean you lose all of the contents. Even if you always use Word and never use any other file format, do yourself a favour and save each version as a .rtf or .txt file as well. Just like a back up, this means that you won’t have lost all of the text if a file goes corrupt. And yes, I do know people who have multiple backup versions of corrupt files: it’s happened to me.

The real problem with word processing packages like Microsoft Word is when files get very large as they tend to do during a PhD. Once you start putting chapters together especially if they include embedded images tables and the like you will find that many word processors start misbehaving. This can be especially frustrating in the [closing phases of your PhD](part4.html) when you're trying to submit it. Especially if there's a tight deadline. However this problem is easily overcome by chunking your thesis into documents equivalent to chapters and then merging PDFs from each chapter.

## Alternatives to conventional word processors

Many academics are turning to latex in order to write their manuscripts as this gives a lot of possibility to insert links, formulae, references, images, etc. in a very simple way. Once you are ready to print a latex document it can be shared very easily as a PDF.

Be aware that the problems may come not with what you want but with how your advisor and or co-authors are comfortable with. I find that commenting on a PDF takes more time than doing the same on a Word document. I don't mind working with latex but I know some advisors are very adverse to having to use this platform. Obviously in order to get the feedback from your advisor you are going to need to use a platform that they are also comfortable with. So please consult them on what they are happy to use. You may also need to instruct them about how to make comments. For example track changes is a function that a lot of people like to use but it doesn't translate back into other platforms seamlessly.

As with many things in this book I urge you to keep an open mind with respect to software. For example if your lab is progressive and already experimenting with new and flexible software platforms then I suggest that you learn with them. The future of computing is certainly likely to be open-source software and so the sooner that you become more familiar with this, the easier you will find it to adapt in future.

Another consideration is how your word processing package is going to work with referencing and citations. Again, your lab may already use some software that you want to integrate with your word processor.

@@

 - Reproducibility
 - Dynamic Content (include statistical code and graphics inside your thesis)
 - Focus on content (not formatting)
 - Numerous output types (including .docx, .html and .pdf)
 - Independent reference manager (via .bib file) and style files (via .csl file)
 - Free software and community support

Importantly, what you want to try to avoid is conducting a major switch between software right at the end of your PhD. Try making such changes when you have more time, like after you’ve finished!

## Naming versions of files

Good file management is an important aspect of a PhD. I like my students to give each chapter of their thesis a nickname or short running title that we both agree on to name the files. An example might be “Swimming_performance_V1.rft”. The reason why I don’t like “Chapter_1.rtf” as a filename is that every student has a chapter 1, so I can’t distinguish between them. In addition, I’m not good at remembering which chapter is chapter 1 and which is chapter 2. Sometimes, towards the end of a thesis, the chapters get their orders switched around anyway, causing even more confusion. Notice that I don’t like having spaces in filenames. This is a hangover from old days of MS-DOS, but is consistently useful today when using R and other coding software. However, some software platforms react oddly with underscore (_), and so you might need to work with a hyphen instead of a space (-).

@@

While you and your advisor can adopt any naming convention you like, you and they need to instantly be able to distinguish between a new version and one that they have already commented on by the file name. No advisor wants to spend time commenting on a chapter and then be told that it was an old version.

## How many versions of each file should you keep?

I would suggest keeping at least one older version of each file to minimise the possibility of losing the content through corrupt files, but you should also be keeping those .rtf or .txt files as well, just in case. Because of my early experience with file corruption, I started to keep all versions of files. This is unnecessary, and I’ve never needed to go back to anything but the last version. Nevertheless, text files take up very little space and it’s not a problem to keep more versions if you want to. More on this in the chapter on [data management](data.html).

### 2021-04-18 2.11-paywall.Rmd
Add files via upload
# some changes after reading Parts I and II

## How to get around the paywall {#paywall}

@@

@@

Many academics also have accounts on academic social media sites (e.g. Academia.edu; ResearchGate.net). You will need to join the network and have your own account, but these house a lot of pdfs that you can’t get elsewhere.

>>>>

# How to get around the paywall {#paywall}

@@

### Go to the intitutional repository of the author

Most university libraries curate an institutional repository of pdfs of published final manuscripts accepted for publication, so-called Green Open Access copies. These are identical to the printed versions save for the typesetting and proof reading that is done after this accepted copy.

@@

Many academics also have accounts on academic social media sites (e.g. Academia.edu; ResearchGate.net). You will need to join the network (using an email address from an academic institution) and have your own account, but these house a lot of pdfs that you can’t get elsewhere.

### 2021-04-18 2.13-style.Rmd
Add files via upload
# some changes after reading Parts I and II

## Writing style {#style}

@@

During my PhD I read a lot. My study species was the subject of tens of thousands of papers, and I was convinced that I'd find what I'd need somewhere in those musty old reprints and heavy volumes that came from the library shelves. It was easy to believe that the authors of those papers had set great puzzles for me to try to understand their content. The realisation that this is not the best form of communication came as a personal epiphany some years ago.

> “The goal of good writing is straightforward: to make your reader's job as easy as possible” 
>      Kevin W. Plaxco [-@plaxco2010art].

This great quote from a paper by Plaxco [-@plaxco2010art] cuts to the heart of why it’s essential to write well. As an editor, I once experienced an author who thought the opposite, and for me this was an epiphany. The submitted manuscript was an impenetrable mess. I could tell that there was good work in there, but as an editor I felt that the information had been made so obscure by the authors that my readers were unlikely to get much out of it. The author was a colleague, and so I decided to phone him and chat about the need for much greater clarity in his manuscript. Experience with email has taught me that they can often be taken in the wrong way; usually in the worst possible way.

The response I got surprised me. The author recognised that his text was dense and was unapologetic. “Let the reader work to make sense of my data,” he said. My epiphany came not because this was a totally alien concept, but because during my PhD days this was exactly how I had thought scientific writing should be done. I had spent so many hours slogging through dense and dreary papers by well-respected figures. The reward when I finally understood what it was that they were trying to communicate felt so great that I believed the puzzle they had set me was what I should set my own readers. Happily, my advisor didn’t feel the same way, but for years I continued to believe that a paper should be a decent puzzle for my readers to crack. This might seem a crazy concept, but it’s not an uncommon problem. For a great take on this in Chemistry, see this great article by Murray [-@murray2011skillful].

This has led to a general perception that "academics stink at writing" - as Steven Pinker put it. But, I think that Pinker is outdated. Certainly, as Plaxco [-@plaxco2010art] has it, there are a lot of ways that academics can improve their writing, for each other. But the belief that all academic writing is going to be accessible to the general public is far fetched. This is because there are a lot of terms when writing journal articles that are there for precision and brevity. These terms will exclude a larger audience, especially when they get so dense that looking them up is simply too arduous. What Plaxco [-@plaxco2010art] urges us to do is to make the rest of the text as accessible and easy as possible.

Any good writing is an artform, and that includes good academic writing. You can't expect to be the best until you've spent some time at it.

@@

### Don’t bamboozle

It’s easy to use jargon. The whole point of jargon is to convey a (usually complex) idea in a short amount of space. Using a word (or two) instead of using several sentences clearly has some advantages. However, there is such a thing as too much jargon. Simply put, it’s unnecessary to use jargon when you can use plain English in the same amount of space. My old tutor at Liverpool University, the late, great Brian Moss, shared the following example of too much jargon [@bump2009large], when plain English would have been much shorter. Do yourself a favor and read the abstract of Bump et al [-@bump2009large]. The fact that it gave Moss a chance to write about moose wasn’t lost on anyone (see Figure \@ref(fig:moss))!

@@

>>>>

# Writing style {#style}

@@

During my PhD I read a lot. My study species was the subject of tens of thousands of papers, and I was convinced that I'd find what I'd needed somewhere in those musty old reprints and heavy volumes that came from the library shelves. It was easy to believe that the authors of those papers had set great puzzles for me to try to understand their content. The realisation that this is not the best form of communication came as a personal epiphany some years ago.

> “The goal of good writing is straightforward: to make your reader's job as easy as possible” 
> ---  Kevin W. Plaxco [-@plaxco2010art].

This great quote from a paper by Plaxco [-@plaxco2010art], who cuts to the heart of why it’s essential to write well. As an editor, I once experienced an author who thought the opposite, and for me this was the epiphany. The submitted manuscript was an impenetrable mess. I could tell that there was good work in there, but as an editor I felt that the information had been made so obscure by the authors that my readers were unlikely to get much out of it. The author was a colleague, and so I decided to phone him and chat about the need for much greater clarity in his manuscript. Experience with email has taught me that they can often be taken in the wrong way; usually in the worst possible way.

The response I got surprised me. The author recognised that his text was dense and was unapologetic. “Let the reader work to make sense of my data,” he said. My epiphany came not because this was a totally alien concept, but because during my PhD days this was exactly how I had thought scientific writing should be done. I had spent so many hours slogging through dense and dreary papers by well-respected figures. The reward came when I finally understood what it was that they were trying to communicate. This felt so great that I believed the puzzle they had set me was what I should set my own readers. Happily, my advisor didn’t feel the same way, but for years I continued to believe that a paper should be a decent puzzle for readers to crack. This might seem a crazy concept, but it’s not an uncommon problem. For a great take on this in Chemistry, see this article by Murray [-@murray2011skillful].

The preponderance of inpenetrable guff has led to a general perception that "academics stink at writing" - as Steven Pinker put it [@pinker2014why]. But, I think that Pinker [-@pinker2014why] is outdated. Certainly, as Plaxco [-@plaxco2010art] has it, there are a lot of ways that academics can improve their writing, for each other. But the belief that all academic writing is going to be accessible to the general public is far fetched. This is because there are a lot of terms when writing journal articles that are there for precision and brevity. These terms will exclude a larger audience, especially when they get so dense that looking them up is simply too arduous. What Plaxco [-@plaxco2010art] urges us to do is to make the rest of the text as accessible and easy as possible.

Any good writing is an artform, and that includes good academic writing. You can't expect to be the best until you've spent some time at it. If you consider a portfolio of your completed written work as 'your experience', you will get some insight into exactly how small this is, even once you've finished your PhD.

@@

## Don’t bamboozle

It’s easy to use jargon. The whole point of jargon is to convey a (usually complex) idea in a short amount of space. Using a word (or two) instead of using several sentences clearly has some advantages. However, there is such a thing as too much jargon. Simply put, it’s unnecessary to use jargon when you can use plain English in the same amount of space. My old advisor at Liverpool University, the late, great Brian Moss, shared the following example of too much jargon by Bump et al [@bump2009large], when plain English would have been much shorter. Do yourself a favor and read the abstract of Bump et al [-@bump2009large]. The fact that it gave Moss a chance to write about moose wasn’t lost on anyone (see Figure \@ref(fig:moss))!

@@

Whatever your writing style, you need to be conciious of your ultimate aim in writing: to communicate with your reader. Making your text as easy to read as possible [as @plaxco2010art put it] isn't easy, but it will be rewarding. The reward you will get is in willing readers to the message that you want to convey. This is something that will stand you in good sted no matter what career you have after your PhD.

### 2021-04-18 2.14-voice.Rmd
Add files via upload
# some changes after reading Parts I and II

## When is it possible to retain your own voice when writing? {#voice}

Should one use “I” or “we” when writing a scientific paper? Although there was a tradition not to use first-person pronouns when writing scientific articles, this has fallen by the wayside in recent years as the use of I or we makes writing simpler for writers and readers. Use “I” if you are the sole author (or in your thesis) and “we” for two or more authors.

You will regularly see “I” or “we” being used in the last paragraph of the introduction and the first paragraph of the discussion. It’s also quite common to use this in the methods section. The reason why these sections have “I” or “we” is quite obvious, as it allows you to place your own voice on actions and decisions. After all, it is your aim and hypothesis for your study, so you should own it. Using first-person pronouns also allow you to be more concise, and there are some great examples here where you can see what happens when you don’t use them.

@@

### Does writing “I” or “we” help people to understand who you are?

@@

### The Vonnegut advice

@@

> "Our audience requires us to be sympathetic and patient teachers, ever willing to simplify and clarify - whereas we would rather soar high above the crowd, singing like nightingales."
> Kurt Vonnegut

>>>>

# When is it possible to retain your own voice when writing? {#voice}

Should one use “I” or “we” when writing a scientific paper? Although there was a tradition not to use first-person pronouns when writing scientific articles, this has fallen by the wayside in recent years, because the use of I or we makes writing simpler for writers and easier to understand for readers. Use “I” if you are the sole author (or in your thesis) and “we” for two or more authors.

You will regularly see “I” or “we” being used in the last paragraph of the introduction and the first paragraph of the discussion. It’s also quite common to use this in the methods section. The reason why these sections have “I” or “we” is quite obvious, as it allows you to place your own voice on your actions and decisions. After all, it is your aim and hypothesis for your study, so you should own it. Using first-person pronouns also allows you to be more concise, and there are some great examples given by Kurt Vonnegut [-@vonnegut1980how] where you can see what happens when you don’t use them.

@@

## Does writing “I” or “we” help people to understand who you are?

@@

## The Vonnegut advice

@@

> "Our audience requires us to be sympathetic and patient teachers, ever willing to simplify and clarify - whereas we would rather soar high above the crowd, singing like nightingales."
> 
> --- Kurt Vonnegut [-@vonnegut1980how]

### 2021-04-18 2.15-concise.Rmd
Add files via upload
# some changes after reading Parts I and II

## Writing concisely {#concise}

So far, I have put an emphasis on a writing formula in order to get started writing. Many students have a problem in knowing where and how to get started with what seems to be a daunting prospect: writing a proposal or writing the entire thesis. I still feel that getting started quickly and efficiently with some confidence is one of your most important steps. But once you've made it through your first draft, you'll begin to find that you need to start refining your writing. That is saying what you need to say in less words and space.

In an ideal world, what goes into your thesis will be of the same quality as the accepted manuscript. The reality is that the thesis is often more wordy, with the result that reviewers and/or editors will ask you to be more concise. How can we achieve this?

Someone else has written a great essay with 10 top rules to help you be concise. The essay is Open Access, and there's no need for me to repeat the words of Scott Hotaling [-@hotaling2020simple]. However, I am going to write some comments of my own on his rules. There are some other posts that are also worth a second look along with this one: see here and here.

Here, the rules are Scott's, the comments are mine:

### Rule 1. Take writing seriously

It is likely, that like me, you chose to do science at school instead of useful subjects such as English and/or other languages. It might have also occurred to you that it was a really bad idea not to have done more earnest work in languages now that you are expected to write like a professional. Certainly, if you did pay more attention in languages in school, you may not be suffering with the rest of us now. Personally, I never received tuition in the grammatical construction of English. Thus I'm left in the dark most of the time about exactly what is wrong. Instead, I simply try to rework a sentence into what I know is correct [and I often don't achieve that]. The lesson then is that should you come across an opportunity to learn more about writing, do take it. 
Although we talk about writing science here, the grammatical rules of writing (in English) are fairly universal (with some local exceptions), so do practice your writing. Maybe write a letter (or long email) instead of making a call sometimes. It'll also help you to read (see here). If you are a keen reader, then you could do worse than reading some Kurt Vonnegut. Scott provided this link to an article I hadn't seen before. Go on, give it a read!

### Rule 2: Identify and stick to your message

@@

### Rule 3: Get to the point

@@

### Rule 4: Keep your Methods and Results contained

@@

### Rule 5: Do not repeat yourself (too often)

@@

Copy and pasting is very easy, and a good way of suddenly producing vast quantities of text. But repeated text is quickly recognised by the reader and appears very boring and cumbersome. This is likely to happen in the Methods and Results sections. If you find yourself deciding that you can simply cut and paste this paragraph while changing the variable names and the numbers, then you are wrong. Don't ever do it.

Also, please don't cut and paste sections of methods from one part of your thesis to another. Just don't do it.

People also have a tendency to build the abstract by copying and pasting text from the main sections. Don't do it. The reader will quickly see the repetitions and become bored. Similarly, the conclusion/summary section is also often copied from lines above. If it's not worth writing again, then it wasn't worth writing before. Make it fresh, and keep it interesting!

### Rule 6: Avoid unnecessary or inefficient “lead‐ins”

Interesting. I'd say that you need some practice and a critical eye to spot these kinds of errors. Having your advisor help or someone who has edited a lot of text. Scott's got some great examples, and so I'll let you peruse them here.

### Rule 7: Use first‐person, active voice

For some reason, most students avoid this at all cost. There appears to be the idea that saying "I" or "we" isn't correct for scientific writing. In fact, it can be the easiest way to avoid the passive voice. I've written more on this elsewhere... see here.

### Rule 8: Remove unnecessary words

@@

### Rule 9: Simplify your language

This is always a good idea. Making three short sentences instead of one very long one is much better. It also helps you avoid complicated grammatical clauses.

### Rule 10: Seek and embrace feedback

I spend a lot of my time reading and commenting on the writing of my students. Every comment is made to improve the document. I do get mad when I find that this help is ignored without an indication why. My suggestions or comments are simply deleted. Certain students become very hard work when they don't even explain why they refuse to change something. You should have noted by now that it's all about being flexible, and being stubborn about anything you've written isn't going to work. Reading the work of these students quickly becomes a chore.

### Rule 11: Read it yourself

Here's a rule that Scott didn't have. Perhaps it should have been inserted higher up the list. As I've said elsewhere (here for example), it would help you to have your work read through by one of your colleagues before you give it to your advisor. But above all, you must be prepared to read __all of your work__ yourself. Not just the first time you write it, but for **every** version. Please don't ever expect your advisor to read something through when you can't be bothered to read it yourself.

Being concise is a great way to write well, and the abstract is a case in point.

Please don't forget to read Scott's 10 rules as written by Scott [@hotaling2020simple].

>>>>

# Writing concisely {#concise}

In this book, I have put an emphasis on a [writing formula](formula.html) in order to get started writing. Many students have a problem in knowing where and how to get started with what seems to be a daunting prospect: writing a proposal or writing the entire thesis. I still feel that getting started quickly and efficiently with some confidence is one of your most important first steps. But once you've made it through your first draft, you'll begin to find that you need to start refining your writing. That is saying what you need to say as much in less words and space. A bit like constantly rewriting your Tweets to get the same information into 140 characters (yes, it's been increased to 280 but that's not as concise).

In an ideal world, what goes into your thesis will be of the same quality as the accepted manuscript. The reality is that the thesis is often more wordy, with the result that reviewers and/or editors will ask you to be more concise. How can you achieve this?

Scott Hotaling has written a great essay with 10 top rules to help you be concise with your own writing. The essay is Open Access, and there's no need for me to repeat Scott's [-@hotaling2020simple] words. However, I am going to write some comments of my own on his rules.

## Rule 1. Take writing seriously

It is likely, that like me, you chose to do science at school instead of useful subjects such as English and/or other languages. It might have also occurred to you that it was a really bad idea not to have done more earnest work in languages now that you are expected to write like a professional. Certainly, if you did pay more attention in languages in school, you may not be suffering with the rest of us now. Personally, I never received tuition in the grammatical construction of English (you may know this already). Thus I'm left in the dark most of the time about exactly what is wrong. Instead, I simply try to rework a sentence into what I know is correct (and I often don't achieve that). The lesson then is that should you come across an opportunity to learn more about writing, do take it.

Use writing professional emails (to your advisor, co-authors or collaborators) as an opportunity to write concise messages that communicate efficiently and precisely the aim of your message. As a PhD student, you should consider yourself a professional and these communications to colleagues will be received with more import if you can write concisely. A personal plea from me would be to ask that you don't start your email by enquiring after the health of your recipients. The fashion to to this has turned into a surge in recent years (the pandemic notwithstanding), and the reality is that this are simply superfluous and gratuitous words. You certainly don't expect them to write back starting with a paragraph about their health. Save this kind of correspondence for your friends and relatives, who may well appreciate you enquiring about their health. When your advisor, co-authors or collaborators become friends, you can move to a less formal style, but I'd argue that you always want to keep your email communication concise.

Although we talk about writing science here, the grammatical rules of writing (in English) are fairly universal (with some local exceptions), so do practice your writing. Maybe write a letter (or long email) instead of making a call sometimes. It'll also help you to [read critically](critical.html). If you are a keen reader, then you could do worse than reading some Kurt Vonnegut [-@vonnegut1980how].

## Rule 2: Identify and stick to your message

@@

## Rule 3: Get to the point

@@

## Rule 4: Keep your Methods and Results contained

@@

## Rule 5: Do not repeat yourself (too often)

@@

Copy and pasting is very easy, and a good way of suddenly producing vast quantities of text (Figure \@ref(fig:plagiarism)). But repeated text is quickly recognised by the reader and appears very boring and cumbersome. This is likely to happen in the Methods and Results sections. If you find yourself deciding that you can simply cut and paste this paragraph while changing the variable names and the numbers, then you are wrong. Don't ever do it.

Also, please don't cut and paste sections of [methods](methods.html) from one part of your thesis to another. Just don't do it.

People also have a tendency to build the [abstract](abstract.html) by copying and pasting text from the main sections. Don't do it. The reader will quickly see the repetitions and become bored. Similarly, the [conclusion/summary](discussion.html) section is also often copied from lines above. If it's not worth writing again, then it wasn't worth writing before. Make it fresh, and keep it interesting!

## Rule 6: Avoid unnecessary or inefficient “lead‐ins”

Interesting. I'd say that you need some practice and a critical eye to spot these kinds of errors. Having your advisor help or someone who has edited a lot of text. Scott's got some great examples [@hotaling2020simple].

## Rule 7: Use first‐person, active voice

For some reason, most students avoid this at all cost. There appears to be the idea that saying "I" or "we" isn't correct for scientific writing. In fact, it can be the easiest way to avoid the passive voice. There are other reasons why [using the first person](voice.html) is helpful.

## Rule 8: Remove unnecessary words

@@

## Rule 9: Simplify your language

This is always a good idea (see Figure \@ref(fig:moss)). Making three short sentences instead of one very long one is much better. It also helps you avoid complicated grammatical clauses.

## Rule 10: Seek and embrace feedback

I spend a lot of my time reading and commenting on the writing of my students. Every comment is made to improve the document. I do get mad when I find that this help is ignored without an indication why. My suggestions or comments are simply deleted. Certain students become very hard work when they don't even explain why they refuse to change something. Reading the work of these students quickly becomes a chore. You should have noted by now that it's all about being flexible, and being stubborn about anything you've written isn't going to work.

## Rule 11: Read it yourself

Here's a rule that Scott didn't have. Perhaps it should have been inserted higher up the list. It would help you to have your work read through by one of your colleagues before you [give it to your advisor](goodenough.html). But above all, you must be prepared to read __all of your work__ yourself, and be prepared to work on it when you read. Edit and improve it on every read. Not just the first time you write it, but for **every** version. Please don't ever expect your advisor to read something through when you can't be bothered to read it yourself.

Being concise is a great way to write well, and the [abstract](abstract.html) is a case in point. Please don't forget to read Scott's 10 rules as written by Scott [@hotaling2020simple].

### 2021-04-18 2.16-foreign.Rmd
Add files via upload
# some changes after reading Parts I and II

## Writing a PhD if English is not your first language {#foreign}

Reading this book you will learn of the many biases in biological sciences and how you need to be aware of them in order to properly interpret your work. Be aware also that biological sciences is biased in where studies take place, and who publishes them [@culumber2019widespread; @nunez2021monolingual]. What we really need then are more studies from less studied parts of the world, and especially we need these to be researched and published by people from those countries. Why then do we insist that science be written in English?

@@

### Message from an editor

@@

### Must a ‘native English speaker’ check your work?

@@

### So if not a native English speaker, who should do it?

@@

### But surely my native tongue can be used for something?

>>>>

# Writing a PhD if English is not your first language {#foreign}

Reading this book you will learn of some biases in biological sciences and how you need to be aware of them in order to properly interpret your work. Be aware also that biological sciences is biased in where studies take place, and who publishes them [@culumber2019widespread; @nunez2021monolingual]. What we really need then are more studies from less studied parts of the world, and especially we need these to be researched and published by people from those countries. Why then do we insist that science be written in English?

@@

## Message from an editor

@@

## Must a ‘native English speaker’ check your work?

@@

## So if not a native English speaker, who should do it?

@@

## But surely my native tongue can be used for something?
